---
title: "The bn-data-v1 JSON Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The bn-data-v1 JSON Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(abn)
library(jsonlite)
```

# Purpose

This vignette specifies `bn-data-v1`, the columnar JSON format for exchanging
raw data between Bayesian-network tools. The format is independent of `abn`;
`abn` provides one import/export adapter.

The generic core contains both observations and the information needed to
interpret them. Tool-specific details belong under namespaced metadata
extensions. For the fitted-network format, see `json-format.Rmd`; for the
mapping to native `abnFit` objects, see `fitabn-json-specification.Rmd`.

# Minimal structure

```json
{
  "metadata": {
    "schema_version": "bn-data-v1",
    "issuer": "abn::export_abnData",
    "label": "example dataset",
    "extensions": {}
  },
  "variables": [
    {
      "name": "height",
      "type": "continuous",
      "role": "node",
      "distribution": "gaussian"
    },
    {
      "name": "status",
      "type": "categorical",
      "role": "node",
      "distribution": "binomial",
      "states": [{"label": "no"}, {"label": "yes"}]
    },
    {
      "name": "subject",
      "type": "identifier",
      "role": "group"
    }
  ],
  "data": {
    "orientation": "columnar",
    "columns": {
      "height": [1.72, 1.81, 1.65],
      "status": ["yes", "no", "yes"],
      "subject": [1, 1, 2]
    }
  },
  "missing_values": {
    "null": null,
    "nan": "NaN",
    "positive_infinity": "Infinity",
    "negative_infinity": "-Infinity"
  }
}
```

The canonical data representation is columnar. Every column must have the same
number of observations. Column names in `data.columns` must correspond to the
declared variable names.

# Generic core

## Metadata

`metadata.schema_version` identifies the format. `metadata.issuer` identifies
the producing tool and operation. `metadata.label` is an optional human-readable
document label. It is not a data column and does not affect interpretation.

Unknown metadata fields and extension namespaces may be ignored by consumers.

## Variables

Each entry in `variables` describes one data column:

| Field | Meaning |
|---|---|
| `name` | Column name and key in `data.columns` |
| `type` | Generic value type, such as `continuous`, `categorical`, `count`, or `identifier` |
| `role` | Use in a workflow, such as `node`, `group`, `weight`, or `ignored` |
| `distribution` | Statistical distribution when the column is a model node |
| `states` | Ordered categorical state declarations |

Distribution and state information belongs to the core because another tool
needs it to interpret the observations and construct a Bayesian-network model.

## Data

The `data` object has:

```json
{
  "orientation": "columnar",
  "columns": {
    "node_a": [1, 2, 3],
    "node_b": ["low", "high", "low"]
  }
}
```

Values are stored separately from their interpretation. Missing values and
special numeric values are represented according to `missing_values` and the
declared column type.

## Missing values

JSON has no native representation for `NaN` or infinite values. A document may
declare explicit tokens for them:

```json
"missing_values": {
  "null": null,
  "nan": "NaN",
  "positive_infinity": "Infinity",
  "negative_infinity": "-Infinity"
}
```

Consumers must not silently reinterpret a declared token as an ordinary
categorical label. Missingness and special numeric values should be restored
according to the column's declared type and encoding.

# Issuer-specific extensions

Extensions must not duplicate generic column properties. They may add
information that is useful to one producer or consumer:

```json
{
  "metadata": {
    "extensions": {
      "abn": {
        "extra": {
          "variables": {
            "subject": {
              "r_class": "factor",
              "original_column_position": 3
            }
          }
        }
      }
    }
  }
}
```

The generic `variables` entry remains the source of truth for `name`, `type`,
`role`, `distribution`, and `states`. An extension that repeats one of those
values is malformed: equal duplicates should warn and conflicting values should
error in strict mode.

# Exporting `abn` data

`export_abnData()` converts an R data frame and a named distribution list into
the portable `bn-data-v1` document. The names of `data.dists` identify model
node columns and their values identify the node distributions.

```{r export-data}
data(g2b2c_data)

data_json <- export_abnData(
  data.df = g2b2c_data,
  data.dists = list(
    G1 = "gaussian",
    B1 = "binomial",
    B2 = "binomial",
    C = "multinomial",
    G2 = "gaussian"
  )
)

cat(data_json)
```

The result can be written directly to a file when supported by the adapter's
file argument, or written explicitly:

```{r write-data, eval = FALSE}
writeLines(data_json, "network-data.json")
```

# Importing into `abn`

The same document can be imported into an R-native data representation:

```{r import-data}
restored <- import_abnData(json = data_json)
str(restored$data.df)
restored$data.dists
```

The returned `data.df` and `data.dists` can be supplied to `buildScoreCache()`
or other `abn` workflows. R-specific reconstruction details, such as factor
classes or row names, may be carried in the ABN extension but are not required
by other Bayesian-network tools.

# Validation

Consumers should validate:

1. `metadata.schema_version` is supported and equals `bn-data-v1`.
2. `metadata.issuer` is present when provenance is required.
3. Variable names are unique and match the column names.
4. All data columns have equal lengths.
5. Declared states are unique and values obey their encoding.
6. Distributions are compatible with variable types and roles.
7. Missing-value tokens are unambiguous for the declared column type.
8. Extensions do not duplicate generic column values.

Unknown extension properties may be ignored by consumers that do not understand
them. Missing extension references are non-fatal but should produce warnings.
