---
title: "The bn-data JSON Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The bn-data JSON Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(abn)
library(jsonlite)
```

# Purpose

This vignette specifies `bn-data`, the columnar JSON format for exchanging
raw data between Bayesian-network tools. The format is independent of `abn`;
`abn` provides one import/export adapter.

Values are stored separately from their interpretation: `data` holds one
array of scalar values per column, `metadata` describes how to read them.
For the fitted-network format, see `vignette("json-format", package = "abn")`;
for the mapping to native `abnFit` objects, see
`vignette("fitabn-json-specification", package = "abn")`.

The normative schema is `inst/schemas/bn-data.schema.json`.

# Minimal structure

```json
{
  "metadata": {
    "schema_version": "bn-data",
    "issuer": "abn::export_abnData",
    "orientation": "columnar",
    "column_order": ["height", "status", "subject"],
    "missing_values": {
      "missing": null,
      "not_a_number": "NaN",
      "positive_infinity": "Infinity",
      "negative_infinity": "-Infinity"
    },
    "columns": [
      {"name": "height", "role": "node", "semantic_type": "continuous",
       "model_distribution": "gaussian", "value_encoding": "number",
       "nullable": false},
      {"name": "status", "role": "node", "semantic_type": "binary",
       "model_distribution": "binomial", "value_encoding": "labels",
       "nullable": false, "categories": ["no", "yes"], "ordered": false},
      {"name": "subject", "role": "group", "semantic_type": "categorical",
       "value_encoding": "labels", "nullable": false}
    ],
    "data_dists": {"height": "gaussian", "status": "binomial"},
    "group_var": "subject",
    "adapters": {"r": {"row_names": ["1", "2", "3"], "columns": {}}}
  },
  "data": {
    "height": [1.72, 1.81, 1.65],
    "status": ["yes", "no", "yes"],
    "subject": ["a", "a", "b"]
  }
}
```

The canonical data representation is columnar. Every column must have the
same number of observations; `metadata.column_order` fixes the order.

# Generic core

## Metadata

`metadata.schema_version` identifies the format. `metadata.issuer` identifies
the producing tool and operation. `metadata.orientation` is `"columnar"`.
`metadata.data_dists` maps model node names to their distributions and
`metadata.group_var` names the grouping column, if any.

Unknown metadata fields and extension namespaces may be ignored by consumers.

## Column descriptors

Each entry of `metadata.columns` describes one data column:

| Field | Meaning |
|---|---|
| `name` | Column name and key in `data` |
| `role` | `node` or `group` |
| `semantic_type` | `continuous`, `count`, `binary`, `categorical`, or `unknown` |
| `model_distribution` | `gaussian`, `binomial`, `poisson`, or `multinomial` for node columns; omitted for group columns |
| `value_encoding` | `labels`, `integer`, `number`, `boolean`, or `string` |
| `nullable` | Whether the column contains missing values |
| `categories` | Declared levels for label-encoded columns |
| `ordered` | Whether the levels of a label-encoded column are ordered |

Distribution and category information belongs to the core because another
tool needs it to interpret the observations and construct a Bayesian-network
model.

## Data

`data` holds one array per column, keyed by column name. JSON has no native
representation for `NaN` or infinite values, so the `missing_values` block
declares the tokens used: JSON `null` for missing, and the strings `"NaN"`,
`"Infinity"`, `"-Infinity"` for the special numerics. Consumers must not
silently reinterpret a declared token as an ordinary categorical label.

## Summary

`export_abnData()` optionally appends a `metadata.summary` block with
recomputed, purely descriptive statistics (`row_count`, per-column counts,
missing counts, and numeric or level summaries). The summary is derived data:
`validate_data_json_summary()` detects stale summaries and
`repair_data_json_summary()` recomputes them.

# R-specific adapter data

`metadata.adapters.r` records R classes, storage types, factor levels and row
names so that `import_abnData()` can reconstruct an R-native data frame
exactly. Other Bayesian-network tools may ignore this block; the generic
column descriptors remain the source of truth for interpretation.

# Exporting `abn` data

`export_abnData()` converts an R data frame and a named distribution list
into the portable `bn-data` document. The names of `data.dists` identify the
model node columns; a `group.var` column is declared with role `group`.

```{r export-data}
data(g2b2c_data)

data_json <- export_abnData(
  data.df = g2b2c_data,
  data.dists = list(
    G1 = "gaussian",
    B1 = "binomial",
    B2 = "binomial",
    C = "multinomial",
    G2 = "gaussian"
  )
)

cat(data_json)
```

A `file` argument writes the document and returns the path invisibly.

# Importing into `abn`

The same document can be imported into an R-native data representation:

```{r import-data}
restored <- import_abnData(json = data_json)
str(restored$data.df)
restored$data.dists
```

The returned `data.df` and `data.dists` can be supplied to
`buildScoreCache()` or `fitAbn()`. A data document can also be attached
directly to a network document with `import_abnFit(json = ..., data = ...)`;
the distributions and factor levels of the two documents are cross-checked
and mismatches are rejected.

# Validation

Consumers should validate:

1. `metadata.schema_version` is supported and equals `bn-data`.
2. `metadata.column_order` names are unique and match both
   `metadata.columns` and the keys of `data`.
3. All data columns have equal lengths.
4. Declared categories are unique and values obey their encoding.
5. `data_dists` and `group_var` are consistent with the column roles.
6. Missing-value tokens are unambiguous for the declared column type.
7. A present `metadata.summary` agrees with the data (or is treated as
   stale).

Unknown metadata fields and extension namespaces may be ignored by
consumers that do not understand them.
