---
title: "Data JSON I/O Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data JSON I/O Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(abn)
library(jsonlite)
```

# Purpose

This vignette specifies the portable JSON format for raw observations used by
Bayesian-network tools. It is a genuine data-exchange format, not an
`abn`-specific dump of an R data frame. `abn` provides one adapter through
`export_abnData()` and `import_abnData()`.

The format is intended for use beyond R. The main payload therefore uses
language-neutral metadata such as `semantic_type`, `categories`, and
`missing_values`. R-specific reconstruction hints may be present, but should
live under an adapter-specific namespace rather than defining the general schema.

The guiding rule is:

- `metadata` describes how to interpret the values.
- `data` contains only the values.
- Metadata should distinguish declarations from derived summaries. Declarations
  such as `schema_version`, `orientation`, `column_order`, `missing_values`, and
  per-column encodings are needed to interpret the data. Derived values such as
  `row_count` and `column_count` can be computed from `data` and should be added
  only when they provide clear value.

The examples use small hand-written documents for clarity and executable
`abn` examples for round trips.

# Top-level structure

The proposed top-level structure separates interpretation from payload:

```json
{
    "metadata": {
      "schema_version": "bn-data-v1",
      "issuer": "abn::export_abnData",
      "orientation": "columnar",
      "column_order": ["G1", "B1", "B2", "C", "G2"],
      "missing_values": {
      "missing": null,
      "not_a_number": "NaN",
      "positive_infinity": "Infinity",
      "negative_infinity": "-Infinity"
    },
    "columns": []
  },
  "data": {}
}
```

The `data` object should not contain type descriptors, roles, factor levels, or
other interpretation metadata. Those belong in `metadata`.

`row_count` and `column_count` are intentionally not part of the minimal example.
For inline JSON data they are derived from the payload (`length(data[[column]])`
and `length(column_order)`). If a document includes them under
`metadata.summary`, importers must validate them against the actual payload and
treat mismatches as errors.

# Metadata

## General metadata

General metadata applies to the whole table:

| Field | Meaning |
| --- | --- |
| `schema_version` | Version of this data JSON structure. |
| `orientation` | Currently expected to be `columnar`. |
| `column_order` | Authoritative order of columns. |
| `missing_values` | Declared tokens for missing and special numeric values. |
| `columns` | Per-column metadata entries. |
| `issuer` | Tool and operation that produced the document. |
| `configs` | Optional settings relevant to interpretation or reproduction. |
| `extensions` | Optional namespaced issuer-specific annotations. |
| `adapters` | Optional language/tool-specific reconstruction hints. |

The `column_order` field is important for `abn`: many data validation paths
expect `data.df` and `data.dists` to have matching names in the same order.
Although a column count can be derived from `column_order`, the order itself is
not redundant because JSON object key order should not be treated as semantic.

## Derived summaries

Derived summaries may be useful for large or externally referenced data, but they
are dangerous if they can get out of sync with the payload. The recommended v1
schema keeps summaries optional for inline data. When present, derived values
belong under `metadata.summary`:

```json
{
  "metadata": {
    "summary": {
      "row_count": 1000,
      "column_count": 5,
      "missing_count": 0,
      "columns": {
        "G1": {"missing_count": 0, "unique_count": 1000}
      }
    }
  }
}
```

Such summaries should be treated as assertions: importers must recompute and
validate them whenever the full `data` payload is available.

## Issuer-specific extensions

Issuer-specific annotations belong under `metadata.extensions` and should be
namespaced by issuer. They can annotate real columns without changing the
portable column definition:

```json
{
  "metadata": {
    "issuer": "some-tool::export-data",
    "extensions": {
      "some-tool": {
        "columns": {
          "C": {"display_color": "red"}
        }
      }
    }
  }
}
```

An extension that refers to a column absent from `column_order` is non-fatal.
Importers should ignore it and issue a warning. Consumers should also ignore
unknown extension namespaces and properties.

The helper `validate_data_json_summary()` implements this policy. A missing
summary is valid because it can be recomputed from inline data. A stale summary
errors by default, can be returned as diagnostics, or can be repaired by replacing
`metadata.summary` with recomputed values:

```{r}
data_json <- list(
  metadata = list(
    schema_version = "bn-data-v1",
    orientation = "columnar",
    column_order = c("G1", "B1"),
    columns = list(
      list(name = "G1", role = "node", semantic_type = "continuous"),
      list(name = "B1", role = "node", semantic_type = "binary")
    )
  ),
  data = list(
    G1 = list(0.1, -0.4, 1.2),
    B1 = list("0", NULL, "1")
  )
)

validate_data_json_summary(data_json, on_error = "return")

with_summary <- repair_data_json_summary(data_json)
with_summary$metadata$summary$row_count

with_summary$metadata$summary$row_count <- 99
validate_data_json_summary(with_summary, on_error = "return")

repaired <- validate_data_json_summary(
  with_summary,
  on_error = "return",
  repair = TRUE
)
repaired$object$metadata$summary$row_count
```

## Column metadata

Each column metadata entry should use language-neutral names where possible:

```json
{
  "name": "C",
  "role": "node",
  "semantic_type": "categorical",
  "model_distribution": "multinomial",
  "value_encoding": "labels",
  "categories": ["1", "2", "3"],
  "ordered": false,
  "nullable": false
}
```

Recommended column fields:

| Field | Meaning |
| --- | --- |
| `name` | Column name. |
| `role` | `node`, `group`, `ignored`, or `metadata`. |
| `semantic_type` | General type such as `continuous`, `binary`, `count`, or `categorical`. |
| `model_distribution` | ABN distribution for node columns: `gaussian`, `binomial`, `poisson`, `multinomial`, or `null`. |
| `value_encoding` | How values are encoded, for example `number`, `integer`, `labels`, or `boolean`. |
| `categories` | Allowed category labels for categorical/binary columns. |
| `ordered` | Whether categories are ordered. |
| `nullable` | Whether the column contains missing values. |

Grouping variables should be represented as `role = "group"`. They may be
categorical, but they are not DAG nodes and usually do not appear in
`data.dists`.

## R adapter metadata

R-specific details should be optional and scoped under `adapters.r`:

```json
{
  "metadata": {
    "adapters": {
      "r": {
        "row_names": null,
        "columns": {
          "C": {
            "class": ["factor"],
            "levels": ["1", "2", "3"]
          }
        }
      }
    }
  }
}
```

This keeps the general format portable while still allowing exact reconstruction
of an R `data.frame` when needed.

# Data payload

The recommended payload is column-oriented:

```json
{
  "data": {
    "G1": [0.1, -0.4, 1.2],
    "B1": ["0", "1", "0"],
    "B2": ["1", "0", "1"],
    "C": ["1", "2", "1"],
    "G2": [3.4, 1.8, 2.1]
  }
}
```

All arrays must have the same length. The set of keys in `data` must match
`metadata.column_order`, and importers should use `column_order` rather than
relying on JSON object ordering. If `metadata.summary.row_count` is present, it
must match the data-derived row count.

# Example: compact mixed data

The `g2b2c_data` dataset is a compact example with Gaussian, binomial, and
categorical columns:

```{r}
data(g2b2c_data)
str(g2b2c_data)
```

A small data JSON document for the first three rows could look like this:

```json
{
    "metadata": {
      "schema_version": "bn-data-v1",
      "orientation": "columnar",
      "column_order": ["G1", "B1", "B2", "C", "G2"],
    "missing_values": {
      "missing": null,
      "not_a_number": "NaN",
      "positive_infinity": "Infinity",
      "negative_infinity": "-Infinity"
    },
    "columns": [
      {"name": "G1", "role": "node", "semantic_type": "continuous", "model_distribution": "gaussian", "value_encoding": "number", "nullable": false},
      {"name": "B1", "role": "node", "semantic_type": "binary", "model_distribution": "binomial", "value_encoding": "labels", "categories": ["0", "1"], "ordered": false, "nullable": false},
      {"name": "B2", "role": "node", "semantic_type": "binary", "model_distribution": "binomial", "value_encoding": "labels", "categories": ["0", "1"], "ordered": false, "nullable": false},
      {"name": "C", "role": "node", "semantic_type": "categorical", "model_distribution": "multinomial", "value_encoding": "labels", "categories": ["1", "2", "3"], "ordered": false, "nullable": false},
      {"name": "G2", "role": "node", "semantic_type": "continuous", "model_distribution": "gaussian", "value_encoding": "number", "nullable": false}
    ]
  },
  "data": {
    "G1": [0.15, -0.28, 0.52],
    "B1": ["0", "1", "0"],
    "B2": ["1", "0", "1"],
    "C": ["1", "2", "3"],
    "G2": [1.91, 2.44, 1.32]
  }
}
```

The numeric values above are illustrative. Tests should use values generated
from the actual data to avoid documentation drift.

# Example: grouped mixed data

The `g2pbcgrp` dataset adds a grouping variable:

```{r}
data(g2pbcgrp)
str(g2pbcgrp)
```

Here the `group` column is metadata for mixed-effect fitting. It is part of the
data table, but its role is `group`, not `node`:

```json
{
    "metadata": {
      "schema_version": "bn-data-v1",
      "orientation": "columnar",
      "column_order": ["G1", "P", "B", "C", "G2", "group"],
    "columns": [
      {"name": "G1", "role": "node", "semantic_type": "continuous", "model_distribution": "gaussian", "value_encoding": "number", "nullable": false},
      {"name": "P", "role": "node", "semantic_type": "count", "model_distribution": "poisson", "value_encoding": "integer", "nullable": false},
      {"name": "B", "role": "node", "semantic_type": "binary", "model_distribution": "binomial", "value_encoding": "labels", "categories": ["0", "1"], "ordered": false, "nullable": false},
      {"name": "C", "role": "node", "semantic_type": "categorical", "model_distribution": "multinomial", "value_encoding": "labels", "categories": ["1", "2", "3"], "ordered": false, "nullable": false},
      {"name": "G2", "role": "node", "semantic_type": "continuous", "model_distribution": "gaussian", "value_encoding": "number", "nullable": false},
      {"name": "group", "role": "group", "semantic_type": "categorical", "model_distribution": null, "value_encoding": "labels", "categories": ["1", "2"], "ordered": false, "nullable": false}
    ]
  },
  "data": {
    "G1": [0.11, -0.31, 0.44],
    "P": [4, 2, 7],
    "B": ["1", "0", "1"],
    "C": ["2", "1", "3"],
    "G2": [1.1, 2.7, 1.8],
    "group": ["1", "1", "2"]
  }
}
```

When reconstructing an `abn` workflow, `data.dists` should be derived only from
columns with `role = "node"`; the grouping column should be passed separately as
`group.var`.

# Validation rules

An importer should validate at least:

1. `metadata.schema_version` is supported.
2. `metadata.orientation` is `columnar`.
3. `metadata.column_order` and `names(data)` contain the same columns.
4. Every `data[[column]]` has the same length.
5. Every column in `metadata.columns` appears exactly once.
6. Categorical values are declared in `categories`, unless they are a declared
   missing value.
7. Count values are non-negative integers or integer-like numbers.
8. `model_distribution` is present only for `role = "node"` columns.
9. Grouping columns are declared with `role = "group"` and are excluded from
   `data.dists`.
10. Any optional derived summary values are recomputed from `data` and must match.

# Relationship to fitted-model JSON

The data and fitted-network formats are complementary. Data JSON describes
observations and their interpretation; network JSON describes a graph and its
fitted quantities. They may be stored separately and joined using stable
variable or column IDs. A combined document may place both genuine objects in
one document, but model export remains model-only by default:

```json
{
  "variables": [],
  "parameters": [],
  "arcs": [],
  "metadata": {"issuer": "some-tool::export-bundle"},
  "data": {}
}
```

For privacy and file-size reasons, fitted-network export should remain model-only
by default, with data export enabled explicitly.

# Design summary

The format separates values from their interpretation, records provenance with
`metadata.issuer`, and leaves room for namespaced extensions. Stable column names
allow annotations to be exchanged without making tool-specific properties part
of the general data schema.
