---
title: "The fitAbn JSON Property Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The fitAbn JSON Property Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Purpose and scope

The vignette **Interoperable JSON Formats for Bayesian Networks**
(`vignette("json-format", package = "abn")`) defines the generic
`bayesian-network` and `bn-data` JSON formats. This vignette defines the
adapter between the generic network format and the R-specific `abnFit`
object.

The specification has two purposes:

1. document where each relevant native `abnFit` property is represented;
2. drive property-specific import and export routines.

The generic JSON core is the interoperability contract: it alone must fully
describe the model. `metadata.extensions.abn` holds only information without
generic meaning. The executable property specification in
`R/json_property_spec.R` is the single source of truth for the mapping; the
table below is generated from it.

# What is an `abnFit` object?

`fitAbn()` returns an object of class `abnFit`. Its contents depend on the
fitting method and on whether a grouping variable was used. The object
contains a fitted Bayesian network together with the fitted conditional
models for its nodes.

The common network component is `abnDag`, which contains:

* `abnDag$dag`, the directed adjacency matrix (rows are children);
* `abnDag$data.dists`, the distribution assigned to each node;
* `abnDag$data.df`, the transformed data used by the R fitting workflow.

Model JSON represents the first two through the generic `variables`,
`arcs`, and `parameters` blocks. `data.df` is not part of a model-only
network document; raw data has its own `bn-data` format, described in
`vignette("bn-data-json-specification", package = "abn")`.

Beyond the network, fits carry the fitted quantities and diagnostics. Since
version 3.2 this also includes information the fitting functions previously
dropped, all of it part of the generic core:

* `centre`: per gaussian node the `(center, scale)` actually applied
  (`variables[].transform` in JSON);
* `levels`: the factor levels used for binomial and multinomial nodes, in
  the order the fit uses them (`variables[].states`);
* `priors`: the prior hyper-parameters of Bayesian fits (`inference.priors`);
* `group.var`, `grouped.vars` for grouped Bayesian fits (`groups`).

## JSON bookkeeping is not an `abnFit` property

The document uses `_id` values for internal references, and `label`,
`scenario_id` and `data_reference` are document metadata. On import these are
kept in `attr(fit, "abn_json")`, never as fields of the fit, and re-export
reuses them so that export -> import -> export is byte-identical.

# Single source of truth

Every native property has one authoritative value location. Nothing is
written twice; values that `abn` can recompute are *derived* on import
instead of stored (Bayes `coef` and `mse` from `modes`; `mdl` from
`mdlnode`). Per-observation information (`group.ids`) is *data* and is
restored from the data document via `import_abnFit(data = ...)`.

The ABN extension contains exactly two things:

* `parameter_names`: for each parameter, the native element (`field`) and
  name it came from, e.g. `{"parameter": 7, "field": "coef", "name": "b3|b1"}`;
* `nodes`: per-node engine diagnostics for Bayesian fits
  (`used_inla`, `error_code`, `error_code_desc`, `hessian_accuracy`).

# Export and import

Export builds the document from the fit (see `R/export.R`): variables and
states from the DAG, distributions and recorded levels; groups; arcs; then
one parameter per additive term, emitted in the order of the native
containers so that re-export is stable. Parameter values are exact: MLE
standard deviations are squared, Bayes precisions stay precisions
(`scale` records which is which).

Import (`R/import.R`) validates the document structurally, rebuilds the fit
from the generic core alone, then overlays the extension: native names come
from `parameter_names` when present and are derived from the structure
otherwise. Documents from other issuers import without any extension.
Dangling references, duplicate `_id`s, missing baselines, and cyclic arcs
are rejected with informative errors.

# The property table

```{r property-table, echo = FALSE, results = "asis"}
spec <- abn:::abn_json_property_spec()
tab <- cbind(
  data.frame(`abnFit property` = paste0("`", spec$field, "`"),
             check.names = FALSE),
  `fit type` = spec$fit_type,
  `JSON location` = ifelse(is.na(spec$location),
                            paste0("*not stored* (", spec$reason, ")"),
                            paste0("`", spec$location, "`"))
)
knitr::kable(tab, format = "pipe", escape = FALSE)
```

`location` paths use `[]` for per-element placement, e.g.
`parameters[].value` or
`inference.diagnostics.nodes[].log_marginal_likelihood`.

# The parameter model

For a node `Y`, the linear predictor is the sum of its applicable
`intercept` and `coefficient` parameters plus the group random effect:

* a `coefficient` with `parent_state` contributes
  `value * 1[x_parent == parent_state]`;
* a `coefficient` without `parent_state` contributes `value * x_parent`
  (after the parent's `transform`, if any);
* for categorical targets, `target_state` selects which non-baseline state
  the parameter belongs to; the baseline state has no parameters;
* `residual_variance`, `random_variance` and `random_covariance` carry a
  `scale` (`"variance"` for MLE, `"precision"` for Bayes).

Because the model is additive there is exactly one `parent` per coefficient.
A node whose multinomial parents are fully one-hot encoded simply has no
intercept parameter. `value` may be `null` for terms `abn` could not estimate
(aliased by collinearity); such terms contribute nothing.

# Reference

See `vignette("json-format", package = "abn")` for the complete generic
network and raw-data formats and `inst/schemas/bayesian-network.schema.json`
for the normative schema. This vignette specifies only the `abn` adapter
mapping and does not replace the generic JSON format.
