---
title: "Interoperable JSON Formats for Bayesian Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interoperable JSON Formats for Bayesian Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(abn)
library(jsonlite)
```

# Purpose

This vignette specifies the JSON formats used to exchange Bayesian-network data
and fitted networks between tools. The formats are deliberately generic. `abn`
is one producer and consumer of these documents, but the documents are not a
serialization of the internal R `abnFit` object.

There are two complementary formats:

* **Data JSON** stores observations and the metadata needed to interpret them.
* **Network JSON** stores variables, graph structure, model parameters, and
  inference results.

They can be exchanged independently or used together. A model-only network
document does not contain observations by default.

# Metadata and extensions

Both formats use a top-level `metadata` object. Generic fields describe the
document itself; issuer-specific annotations live in `metadata.extensions`.

```json
{
  "metadata": {
    "schema_version": "bayesian-network",
    "issuer": "abn::export_abnFit",
    "issuer_version": "3.1.13",
    "created": "2026-10-07T12:00:00Z",
    "label": "optional descriptive name",
    "scenario_id": "optional run identifier",
    "data_reference": {"uri": "model_data.json", "schema_version": "bn-data"},
    "extensions": {
      "abn": {
        "parameter_names": [
          {"parameter": 4, "field": "coef", "name": "status|height"}
        ],
        "nodes": [
          {"variable": 2, "used_inla": true, "error_code": 0,
           "error_code_desc": "success"}
        ]
      }
    }
  }
}
```

`issuer` identifies the tool and operation that produced the document. Generic
consumers should not need to understand it. `extensions` contains
issuer-specific information, keyed by issuer name; consumers must ignore
extension namespaces they do not know.

Everything with a generic meaning belongs in the core blocks, never in an
extension: the extension holds only information that is meaningless outside
`abn`, such as the native R names of coefficients or engine diagnostics. Each
value is stored exactly once.

`_id` values are document-local join keys. They must be unique within each
object collection, but have no meaning outside the document. References must
use `_id` values, never display names. Unknown extension namespaces and
properties should be ignored by consumers that do not understand them.

# Data JSON

The data format is identified by `metadata.schema_version = "bn-data"`.
The payload is column-oriented and contains values only; interpretation belongs
in metadata.

```json
{
  "metadata": {
    "schema_version": "bn-data",
    "issuer": "abn::export_abnData",
    "orientation": "columnar",
    "column_order": ["height", "status"],
    "missing_values": {
      "missing": null,
      "not_a_number": "NaN",
      "positive_infinity": "Infinity",
      "negative_infinity": "-Infinity"
    },
    "columns": [
      {
        "name": "height",
        "role": "node",
        "semantic_type": "continuous",
        "model_distribution": "gaussian",
        "value_encoding": "number"
      },
      {
        "name": "status",
        "role": "node",
        "semantic_type": "binary",
        "model_distribution": "binomial",
        "value_encoding": "labels",
        "categories": ["no", "yes"]
      }
    ],
    "extensions": {}
  },
  "data": {
    "height": [1.72, 1.81, 1.65],
    "status": ["yes", "no", "yes"]
  }
}
```

Column metadata can describe node, group, ignored, and auxiliary columns.
`model_distribution` is used for node columns and is omitted for grouping
columns. Categories, encodings, ordering, nullability, and declared special
numeric values make the representation portable across languages.

R-specific details such as factor levels and row names may be placed under an
adapter namespace such as `metadata.adapters.r`; they do not define the general
format. Optional derived summaries belong under `metadata.summary` and must be
validated against inline data.

## Using data JSON with `abn`

`abn` can export and import the portable data document as follows:

```{r}
data(g2b2c_data)
data_json <- export_abnData(
  data.df = g2b2c_data,
  data.dists = list(G1 = "gaussian", B1 = "binomial", B2 = "binomial",
                    C = "multinomial", G2 = "gaussian")
)
restored <- import_abnData(json = data_json)
str(restored$data.df)
restored$data.dists
```

The restored `data.df` and `data.dists` can then be supplied to the normal
`abn` structure-learning workflow.

This section is the normative description of the raw-data format. The former
standalone data JSON vignette has been consolidated here so that the two JSON
formats have one authoritative documentation home.

# Network JSON

The network format is identified by `metadata.schema_version =
"bayesian-network"`. Its interoperable core consists of five blocks:

* `variables` declares nodes: generic `type`, `distribution` and `link`, plus
  `states` (with one `baseline` state) for binary and categorical variables.
* `groups` declares the grouping variables of mixed-effects models (possibly
  empty), referencing the variables fitted with a random intercept.
* `arcs` declares directed edges using source and target `_id` values.
* `parameters` declares fitted quantities: one entry per additive term, with
  `kind` (`intercept`, `coefficient`, `residual_variance`, `random_variance`,
  `random_covariance`), its `target` (and `target_state` for categorical
  targets), its single `parent` (and `parent_state` for coded parents), its
  `group`, `value` (possibly `null` for non-estimable terms) and `uncertainty`.
* `inference` contains the fitting `type` (`maximum_likelihood` or `bayesian`),
  the `priors` of Bayesian fits, generic `diagnostics` (network level and per
  node), and posterior `marginals` when available.

The linear predictor of a node is the sum of its applicable `intercept` and
`coefficient` parameters plus the group random effect; a `coefficient` with
`parent_state` contributes `value * 1[x_parent == parent_state]`, otherwise
`value * x_parent` (after the parent's optional `transform`). No coding
convention is needed: a node whose parents are fully one-hot encoded simply
has no intercept parameter.

```json
{
  "metadata": {"schema_version": "bayesian-network", "issuer": "some-tool"},
  "variables": [
    {"_id": 1, "name": "height", "type": "continuous",
     "distribution": "gaussian", "link": "identity"},
    {"_id": 2, "name": "status", "type": "binary",
     "distribution": "bernoulli", "link": "logit",
     "states": [{"_id": 1, "label": "no", "baseline": true},
                {"_id": 2, "label": "yes", "baseline": false}]}
  ],
  "groups": [],
  "arcs": [{"source": 1, "target": 2}],
  "parameters": [
    {"_id": 1, "target": 1, "kind": "intercept", "value": 1.2,
     "uncertainty": {"standard_error": 0.1}},
    {"_id": 2, "target": 1, "kind": "residual_variance",
     "scale": "variance", "value": 0.25},
    {"_id": 3, "target": 2, "kind": "intercept", "value": -0.42,
     "uncertainty": {"standard_error": 0.2}},
    {"_id": 4, "target": 2, "kind": "coefficient", "parent": 1,
     "value": 0.87, "uncertainty": {"standard_error": 0.3}}
  ],
  "inference": {
    "type": "maximum_likelihood",
    "diagnostics": {
      "log_marginal_likelihood": -123.4,
      "nodes": [{"variable": 1, "log_marginal_likelihood": -55.0, "df": 98}]
    }
  }
}
```

Variance-type parameters carry a `scale`: `"variance"` for maximum-likelihood
fits and `"precision"` for Bayesian ones, matching what the fitting engine
natively estimates; consumers convert as needed.

The generic core contains only information needed by an independent
Bayesian-network tool to understand, display, or use the mathematical model.
Engine-specific information that has no generic meaning lives under
`metadata.extensions.abn`: the native `abn` coefficient names and the
per-node INLA diagnostics.

## Always-complete documents

A network document always contains all six blocks (`metadata`, `variables`,
`groups`, `arcs`, `parameters`, `inference`); `groups` may be empty. A document
without the structure would not describe a model, so there is no option to
omit it. Observations never belong to a network document; they are exchanged
through the data format and can be linked via `metadata.data_reference`.

## Using network JSON with `abn`

An `abn` fit can be exchanged through the generic network format:

```{r}
data(ex1.dag.data)
mydists <- list(b1 = "binomial", p1 = "poisson", g1 = "gaussian",
                b2 = "binomial", p2 = "poisson", g2 = "gaussian",
                b3 = "binomial", g3 = "gaussian")
cache <- buildScoreCache(ex1.dag.data, data.dists = mydists,
                          method = "mle", max.parents = 2)
fit <- fitAbn(mostProbable(cache), method = "mle")

network_json <- export_abnFit(fit)
fit_again <- import_abnFit(json = network_json)
```

The import operation is an `abn` adapter: it reconstructs an R-native object
from the generic document. Other Bayesian-network tools can consume the generic
core without reconstructing an `abnFit` object.

# Validation and compatibility

Consumers should validate the following core rules:

1. The schema version is supported and required collections have the right shape.
2. `_id` values are unique within each collection and all references resolve.
3. Arc, parameter, state, and extension references resolve to declared `_id` values.
4. Variable types, states, links, and parameter kinds are compatible.
5. Arcs do not contain self-links or duplicate edges.
6. A network declared as a DAG contains no directed cycles.
7. Data columns have consistent lengths and obey their declared encodings.

Unknown top-level optional fields, metadata fields, extension namespaces, and
extension properties should be ignored unless strict validation is requested.
Extensions must not repeat a generic value; every value belongs to exactly one
location in the document.

# Design summary

The formats separate portable Bayesian-network concepts from implementation
details. `metadata.issuer` provides provenance, `metadata.label` and
`metadata.scenario_id` may provide an optional label and run identifier, and
`metadata.extensions` provides namespaced ABN annotations and engine
diagnostics. Document-local `_id` values are JSON bookkeeping; they make
references and overlays safe to join. `abn` keeps them stable across a
round-trip so that re-export is idempotent.

See `export_abnData()`, `import_abnData()`, `export_abnFit()`, and
`import_abnFit()` for the `abn` adapters.
