---
title: "JSON Format Quick Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JSON Format Quick Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(abn)
```

# JSON Format Quick Reference

This vignette provides a quick reference guide for the JSON format used by `export_abnFit()` and `import_abnFit()` functions in the `abn` package. The JSON format enables round-trip serialization of Bayesian network models, making it ideal for sharing, archiving, and integration with external tools like abnScripts.

## Complete Field Reference

The JSON format contains 10 top-level fields organized into three categories:

### Required Fields (must always be present)

| Field | Type | Description |
|-------|------|-------------|
| `variables` | Array | Definition of all variables in the network (name, id, type, n_categories) |
| `parameters` | Array | Fitted parameters for each node and its parent configurations |
| `arcs` | Array | Directed edges in the network structure (parent_id, child_id) |

### Optional Fields (may be absent or null)

| Field | Type | Description |
|-------|------|-------------|
| `scenario_id` | String or null | User-defined scenario identifier for grouping related models |
| `label` | String or null | Human-readable label or description of the model |
| `method` | String or null | Fitting method used: "bayes", "mle", or null |
| `linkFunctions` | Object or null | Link functions per node (e.g., log, logit, identity) |
| `constraints` | Array or null | Constraints applied during structure learning (prohibited/enforced arcs) |
| `subset_metadata` | Object or null | Metadata about data subsetting (row selection, filtering criteria) |
| `original_model` | Object or null | Reference to the source model for tracing model lineage |
| `original_data_path` | String or null | Path or URI to the original data file used for fitting |

## Common Workflow Examples

### 1. Export a Fitted Model

Export an `abnFit` object after parameter learning:

```{r eval = FALSE}
# After fitting a model with fitAbn()
fit <- fitAbn(...)

# Export to JSON string
json_string <- export_abnFit(fit, format = "json")

# Or export to file
export_abnFit(fit, format = "json", 
              outfile = "my_model.json")
```

### 2. Import and Re-use a Model

Import a previously exported model for prediction or inspection:

```{r eval = FALSE}
# Import from file
fit <- import_abnFit(infile = "my_model.json")

# Or import from JSON string
json_string <- readLines("my_model.json")
fit <- import_abnFit(data_or_string = json_string)
```

### 3. Round-Trip Verification

Export a model and re-import to verify consistency:

```{r eval = FALSE}
# Original model
fit_original <- fitAbn(...)

# Export and re-import
json_data <- export_abnFit(fit_original)
fit_reimported <- import_abnFit(data_or_string = json_data)

# Both models should be equivalent for predictions
```

### 4. Error Handling

The `import_abnFit()` function validates JSON structure and provides informative errors:

```{r eval = FALSE}
tryCatch(
  fit <- import_abnFit("invalid.json"),
  error = function(e) {
    message("Import failed: ", e$message)
    # Handle error - provide feedback to user
  }
)
```

## Variable Definition Structure

Each variable in the `variables` array must include:

```json
{
  "id": "unique_identifier",
  "name": "variable_name",
  "type": "gaussian|binomial|poisson|multinomial",
  "n_categories": 2
}
```

- **id**: Unique identifier used to reference the variable in arcs and parameters
- **name**: Human-readable variable name
- **type**: Data type of the variable
- **n_categories**: For multinomial variables, the number of categories (≥2)

## Parameter Definition Structure

Each parameter represents either node intercepts or parent-conditional parameters:

```json
{
  "id": "unique_identifier",
  "node_id": "variable_id",
  "parent_ids": ["parent_var_id_1", "parent_var_id_2"],
  "coefficient": 0.5,
  "intercept": -0.2
}
```

- **id**: Unique parameter identifier
- **node_id**: ID of the variable this parameter belongs to
- **parent_ids**: Array of parent variable IDs (empty for root nodes)
- **coefficient**: Fitted coefficient value
- **intercept**: Fitted intercept value (may be null for some models)

## Arc Definition Structure

Each arc represents a directed edge from parent to child:

```json
{
  "parent_id": "parent_variable_id",
  "child_id": "child_variable_id"
}
```

- **parent_id**: ID of the parent variable
- **child_id**: ID of the child variable
- Arcs define the network structure; parameters define the relationships

## Quick Reference Tables

### Supported Variable Types

| Type | Support | Use Case |
|------|---------|----------|
| Gaussian | Bayes, MLE | Continuous data (heights, measurements) |
| Binomial | Bayes, MLE | Binary outcomes (yes/no, 0/1) |
| Poisson | Bayes, MLE | Count data (events, occurrences) |
| Multinomial | MLE only | Categorical data with 3+ categories |

### Fitting Methods

| Method | Algorithm | When to Use |
|--------|-----------|------------|
| "bayes" | Bayesian inference | Default; handles small sample sizes well |
| "mle" | Maximum likelihood | Preferred for multinomial variables; larger datasets |

### Link Functions (when applicable)

| Link Function | Applicable To | Interpretation |
|---------------|---------------|-----------------|
| identity | Gaussian | No transformation of linear predictor |
| log | Poisson | Exponential relationship; parameters on log scale |
| logit | Binomial | Odds ratio interpretation; probabilities on logit scale |

## JSON Validation Rules

When importing JSON, the following rules are enforced:

1. **Structure**: Must be a valid JSON object with required fields
2. **Arrays**: `variables`, `parameters`, and `arcs` must be arrays
3. **IDs**: Variable IDs must be unique and consistently referenced
4. **Parent References**: Parent IDs in `arcs` and `parameters` must reference valid variables
5. **Types**: Variable types must be one of: "gaussian", "binomial", "poisson", "multinomial"
6. **Categories**: Multinomial variables must have `n_categories ≥ 2`

## Integration with abnScripts

The JSON format includes fields specifically designed for integration with the abnScripts framework:

- **constraints**: Stores structure learning constraints (prohibited/enforced arcs)
- **subset_metadata**: Tracks data subsetting operations for reproducibility
- **original_model**: Enables model lineage tracking
- **original_data_path**: References the source data for auditing

These fields are optional but recommended when models are part of a larger analysis pipeline.

## See Also

- `export_abnFit()`: Export `abnFit` objects to JSON
- `import_abnFit()`: Import JSON back to `abnFit` objects
- `fitAbn()`: Fit parameters of Bayesian network structures
- `buildScoreCache()`: Score networks for structure learning
