---
title: "Interoperable JSON Formats for Bayesian Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interoperable JSON Formats for Bayesian Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(abn)
library(jsonlite)
```

# Purpose

This vignette specifies the JSON formats used to exchange Bayesian-network data
and fitted networks between tools. The formats are deliberately generic. `abn`
is one producer and consumer of these documents, but the documents are not a
serialization of the internal R `abnFit` object.

There are two complementary formats:

* **Data JSON** stores observations and the metadata needed to interpret them.
* **Network JSON** stores variables, graph structure, model parameters, and
  inference results.

They can be exchanged independently or used together. A model-only network
document does not contain observations by default.

# Metadata and extensions

Both formats use a top-level `metadata` object. Generic fields describe the
document itself; issuer-specific settings and annotations are kept in extensions.

```json
{
  "metadata": {
    "schema_version": "bayesian-network-v1",
    "issuer": "abn::export_abnFit",
    "issuer_version": "3.1.13",
    "configs": {},
    "extensions": {
      "abn": {
        "configs": {},
        "extra": {
          "variables": {
            "2": {"native_name": "status"}
          }
        }
      }
    }
  }
}
```

`issuer` identifies the tool and operation that produced the document. Generic
consumers should not need to understand it. `configs` contains settings that
affect interpretation or reproducibility. `extensions` contains issuer-specific
information attached to genuine objects in the document.

Extension namespaces should identify their issuer. Within an extension, object
annotations are keyed by the corresponding object ID. For example,
`extensions.abn.variables["2"]` annotates the variable with `_id` `2`.
The `extra` object may contain `variables`, `arcs`, `parameters`, or `inference`
collections keyed by the corresponding document-local `_id`. These are overlays:
they contain only issuer-specific properties and must not repeat generic values.

`_id` values are document-local join keys. They must be unique within each
object collection, but have no meaning outside the document and need not be
preserved across exports. References must use `_id` values, never display names.
An extension referring to an ID
that is absent from the genuine object is not fatal: importers should ignore it
and issue a warning. Unknown extension namespaces and properties should likewise
be ignored by consumers that do not understand them.

# Data JSON

The data format is identified by `metadata.schema_version = "bn-data-v1"`.
The payload is column-oriented and contains values only; interpretation belongs
in metadata.

```json
{
  "metadata": {
    "schema_version": "bn-data-v1",
    "issuer": "abn::export_abnData",
    "orientation": "columnar",
    "column_order": ["height", "status"],
    "missing_values": {
      "missing": null,
      "not_a_number": "NaN",
      "positive_infinity": "Infinity",
      "negative_infinity": "-Infinity"
    },
    "columns": [
      {
        "name": "height",
        "role": "node",
        "semantic_type": "continuous",
        "model_distribution": "gaussian",
        "value_encoding": "number"
      },
      {
        "name": "status",
        "role": "node",
        "semantic_type": "binary",
        "model_distribution": "binomial",
        "value_encoding": "labels",
        "categories": ["no", "yes"]
      }
    ],
    "extensions": {}
  },
  "data": {
    "height": [1.72, 1.81, 1.65],
    "status": ["yes", "no", "yes"]
  }
}
```

Column metadata can describe node, group, ignored, and auxiliary columns.
`model_distribution` is used for node columns and is omitted for grouping
columns. Categories, encodings, ordering, nullability, and declared special
numeric values make the representation portable across languages.

R-specific details such as factor levels and row names may be placed under an
adapter namespace such as `metadata.adapters.r`; they do not define the general
format. Optional derived summaries belong under `metadata.summary` and must be
validated against inline data.

## Using data JSON with `abn`

`abn` can export and import the portable data document as follows:

```{r}
data(g2b2c_data)
data_json <- export_abnData(
  data.df = g2b2c_data,
  data.dists = list(G1 = "gaussian", B1 = "binomial", B2 = "binomial",
                    C = "multinomial", G2 = "gaussian")
)
restored <- import_abnData(json = data_json)
str(restored$data.df)
restored$data.dists
```

The restored `data.df` and `data.dists` can then be supplied to the normal
`abn` structure-learning workflow.

This section is the normative description of the raw-data format. The former
standalone data JSON vignette has been consolidated here so that the two JSON
formats have one authoritative documentation home.

# Network JSON

The network format is identified by `metadata.schema_version =
"bayesian-network-v1"`. Its interoperable core consists of four concepts:

* `variables` declares nodes and their types or states.
* `arcs` declares directed edges using source and target `_id` values.
* `parameters` declares fitted quantities and their targets, parents, links, and
  values.
* `inference` contains generic fitting method, estimates, uncertainty, posterior
  results, and diagnostics that are meaningful independently of the producing
  software.

```json
{
  "metadata": {
    "schema_version": "bayesian-network-v1",
    "issuer": "abn::export_abnFit",
    "issuer_version": "3.1.13",
    "configs": {},
    "extensions": {}
  },
  "variables": [
    {"_id": 1, "name": "height", "type": "continuous"},
    {
      "_id": 2, "name": "status", "type": "binary",
      "states": [{"_id": 1, "label": "no"},
                  {"_id": 2, "label": "yes"}]
    }
  ],
  "arcs": [{"source": 1, "target": 2}],
  "parameters": [
    {
      "_id": 1,
      "target": 2,
      "kind": "intercept",
      "link": "logit",
      "value": -0.42
    },
    {
      "_id": 2,
      "target": 2,
      "parents": [1],
      "kind": "coefficient",
      "link": "logit",
      "value": 0.87
    }
  ],
  "inference": {
    "type": "bayesian",
    "estimates": {},
    "uncertainty": {},
    "diagnostics": {}
  }
}
```

The core deliberately uses generic terms such as `type`, `kind`, `target`,
`parents`, and `value`. A tool may add standard uncertainty fields or richer
posterior representations without exposing its internal object layout.

The generic core contains only information needed by an independent
Bayesian-network tool to understand, display, or use the mathematical model.
For example, an `abn` exporter records engine-specific diagnostics under
`metadata.extensions.abn`, while generic inference information remains in
`inference`:

```json
{
  "metadata": {
    "extensions": {
      "abn": {
        "configs": {"compute_fixed": true},
        "diagnostics": {
          "used_inla": {"status": true},
          "error_code": {"status": 0},
          "error_code_desc": {"status": "success"},
          "hessian_accuracy": {"status": 0}
        },
        "extra": {
          "parameters": {
            "3": {"native_name": "status|height"}
          }
        }
      }
    }
  }
}
```

## Network structure and `include_network`

`export_abnFit(..., include_network = TRUE)` includes the DAG representation:
the `variables` and `arcs` collections. This is the default because a fitted
network is normally not interpretable without its graph. With
`include_network = FALSE`, the exporter omits the network structure while still
exporting fitted quantities and their object references. Such a document is
useful when the graph is managed separately, but a consumer must obtain the
referenced variables and arcs from another source.

## Using network JSON with `abn`

An `abn` fit can be exchanged through the generic network format:

```{r}
data(ex1.dag.data)
mydists <- list(b1 = "binomial", p1 = "poisson", g1 = "gaussian",
                b2 = "binomial", p2 = "poisson", g2 = "gaussian",
                b3 = "binomial", g3 = "gaussian")
cache <- buildScoreCache(ex1.dag.data, data.dists = mydists,
                          method = "mle", max.parents = 2)
fit <- fitAbn(mostProbable(cache), method = "mle")

network_json <- export_abnFit(fit, include_network = TRUE)
fit_again <- import_abnFit(json = network_json)
```

The import operation is an `abn` adapter: it reconstructs an R-native object
from the generic document. Other Bayesian-network tools can consume the generic
core without reconstructing an `abnFit` object.

# Validation and compatibility

Consumers should validate the following core rules:

1. The schema version is supported and required collections have the right shape.
2. `_id` values are unique within each collection and all references resolve.
3. Arc, parameter, state, and extension references resolve to declared `_id` values.
4. Variable types, states, links, and parameter kinds are compatible.
5. Arcs do not contain self-links or duplicate edges.
6. A network declared as a DAG contains no directed cycles.
7. Data columns have consistent lengths and obey their declared encodings.

Unknown top-level optional fields, metadata fields, extension namespaces, and
extension properties should be ignored unless strict validation is requested.
The `extra` overlay must not repeat a generic value under `metadata.extensions`;
duplicate values are malformed even when they happen to be equal.
Missing extension `_id` references are specifically non-fatal and should produce warnings.

# Design summary

The formats separate portable Bayesian-network concepts from implementation
details. `metadata.issuer` provides provenance, `metadata.configs.label` may
provide an optional document label, and `metadata.extensions` provides
namespaced ABN annotations and engine diagnostics. Document-local `_id` values
are JSON bookkeeping only; they make overlays safe to join and may be
regenerated by every export.

See `export_abnData()`, `import_abnData()`, `export_abnFit()`, and
`import_abnFit()` for the `abn` adapters.
